Visual diagnostics for constrained optimisation with application to guided tours

### 2020-10-28 paper/paper.Rmd
update paper.Rmd
removed name T for linear projection + minor corrections (up to section 5)

For a given problem we might rely on numeric methods to find the optimal solution, since an analytical approach, requiring a closed-form solution, is often not possible for complex problems.
An iterative approach finds the optimum by progressively improving the current solution. Researchers can then decide when to stop if a desirable accuracy has been achieved. One potential issue with this technique is that it may end up finding a local optimum.

@@

To define projection pursuit, we first need to establish the notation used. Let $\mathbf{X}_{n \times p}$ be the data matrix, with $n$ observations in $p$ dimensions. A d-dimensional projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$, and defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. Define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ (corresponding to an associated projection matrix $\mathbf{A}$) onto an index value $I$. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a projection.

@@

As a general visualisation method, a tour produces animations of high dimensional data via rotations between low dimension planes. Different tour types choose these planes differently, for example, a grand tour[@cook2008grand] selects the planes randomly and a manual tour[@cook1997manual] gradually phases in and out one variable, to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses planes with the aid of projection pursuit. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs the geodesic interpolation between these planes to form a tour path.

@@

where $I_d$ is the d-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be an orthogonal matrices.

Several features of this optimisation are worth noticing. First of all, this is a constraint optimisation problem as the decision variables form the entries of a projection basis, which is required to be orthonormal. It is also likely that the objective function may not be differentiable for a constructed index function and in these cases, gradient-based methods may not work well. Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. Lastly, there is also one computational consideration: the optimisation procedure needs to be fast to compute since the tour animation is played in real-time.

@@

Simulated annealing (`search_better_random`) [@kirkpatrick1983optimization, @bertsimas1993simulated] uses the same sampling process as `search_better` but allow a probabilistic acceptance of a basis with lower index value based on the annealing $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maximum is presented. The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

@cook1995grand used a gradient ascent algorithm on the space of the projection bases. In gradient ascent, one first find the direction for improvement via computing the gradient information. In `search_geodesic`, $2n$ bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter $\delta$. A geodesic is then constructed using the current basis and the one in $2n$ bases with the highest index value. If the neighbourhood parameter $\delta$ is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and works as the searching direction. The next step in gradient ascent is to conduct a line search to find the best improvement along the gradient direction and in `search_geodesic`, this is replaced by optimising the index value along the geodesic direction over an 90 degree angle from $-\pi/4$ to $\pi/4$. The optima $\mathbf{A}_{**}$ is returned for the current iteration if it meets the termination condition on percentage improvement.  The procedure will also terminate if $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. Possible values for `search_better` and `search_better_random` includes `random_search`,  `new_basis`, and `interpolation`. `search_geodesic` has a wider variety that includes `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

Multiple iterators are also  needed to index the data collected at different levels. $t$ is a unique identifier that prescribes the natural ordering of each observation; $j$ is the counter for each search-and-interpolate iteration, which remains the same within one round and has an increment of one once a new round starts. $l$ is the counter for each search/interpolation, which provides the information of how many basis the algorithm has searched before finding one to return. There are other parameters of interest, depends on the particular problem content and they are denoted as *$V_{p}$*. Two most common examples include $V_1 = \text{method}$, which tags the name of the algorithm used, and $V_2 = \text{alpha}$, the neighbourhood parameter that controls the size in sampling candidate bases.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser can't find a better basis and the algorithm terminates. In this notation, final basis found is $A_T$ with the highest index value $I_T$.

@@

The data structure constructed above meets the tidy data principle [@wickham2014tidy] that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by the well-developed packages like dplyr[@dplyr] and ggplot2[@ggplot2].

The construction of diagnostic plots uses the concept of grammar of graphic [@wickham2010layered] in ggplot2. In grammar of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by stacked layers. In the construction of diagnostic plots, there are multiple elements one may wants to emphasize and there's no single plot name that would meet the need, the stacked layers concept, on the other hand, allow information to be overlaid on the each other and one can build the plot from scratch as long as the variables have been stored in a dataset.

## Check how hard the optimiser is working

A primary interest of diagnosing an optimiser is to study how it progressively finds its optimum. Directly plotting the index value across its natural order will cause the graph to be disproportional to the iteration since it usually takes longer for an optimiser to find a better basis towards the end. Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that can provide five points summary of each iteration. Other additional information not presented in the  boxplot and then be added with new layers, for example text information on the number of points can be added at the bottom of each iteration and the position of basis returned by projection pursuit can be highlighted in point. Further, an option to switch between displaying points and boxplot geometry is helpful when the number of observation is small in one iteration and this is achieved via a `cutoff` parameter.

Figure \@ref(fig:toy-search) shows a sample of the plot constructed using `search_better` with different value on the parameter `max_tries`. Comparing the returned index value of each iteration shows that the termination at `max_tries = 25` is not sufficient for `search_better` to explore the parameter space and a value of 500 is preferred over 25 in this context.

@@

(ref:toy-search) A comparison of `search_better` with different parameter value on `max_tries`. A six variable dataset `boa6` is used with `holes` index on a 2D problem. A `max_tries = 500` since it allows the optimiser to find better basis with higher index value at iteration six.

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \@ref(fig:toy-interp) presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decreases pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost cases is a deliberate construction of `search_better_random` where an inferior basis can be accepted in a probability manner so as to avoid trapping in the local maximum.

@@

(ref:toy-pca) 1D projection on the 5 variable dataset `boa5` with two optimisers: `search_better` and `search_geodesic`. The yellow point corresponds to the theoretical best basis [0, 1, 0, 0, 0] with `V2` being the only non-normal variable in the dataset.  The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in `search_geodesic` has been flipped positive to ensure that bases with same projected images but a sign difference are represented by the same point in the plot.

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases $\mathbf{A}_{p \times d}$ lives on the surface of a $p \times d$ dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space (d > 3), an easier alternative is to view the reduced space via some dimension reduction methods, i.e. principal component analysis. To better perceive the the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package `geozoo` and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \@ref(fig:toy-pca) plots the first two principal components of two search paths, one using `search_better` and another uses `search_geodesic`. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in `search_better` is broader, controlled by `alpha` parameter, which is default to 0.5 while the directional search in `search_geodesic`, controlled by `delta` with a default of 0.01, is so tiny that it can barely seen.

@@

(ref:toy-tour) A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row view the space from a direction that is close to the one in PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.

Viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original $p \times d$ dimension enables a more holistic stereoscopical view of the search. To view a high dimensional ($d \ge 3$) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compare to the PCA plot, the animated rotation (tour) display in Figure \@ref(fig:toy-tour) gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and grand tour produces a sequence of angle that view the search from different direction. As an evidence, the last frame in Figure \@ref(fig:toy-tour) is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to one in Figure \@ref(fig:toy-pca).

@@

For a particular index function, the best algorithm to optimise relates to the character of the index and the data. If the index function is smooth and has single maximum, all of the three algorithms introduced above can find the maximum.  When multiple optima are presented, `search_better` may stuck at the local maximum and in the case where the index function is non-smooth, `search_geodesic` may even fail to find the maximum. In this section, examples will be presented  to outline how the diagnostic plots can be used to compare the performance of optimisers in different scenarios.

>>>>

For a given problem we might rely on numerical methods to find the optimal solution, since an analytical approach, requiring a closed-form solution, is often not possible for complex problems.
An iterative approach finds the optimum by progressively improving the current solution. Researchers can then decide when to stop if a desirable accuracy has been achieved. One potential issue with this technique is that it may end up finding a local optimum.

@@

To define projection pursuit, we first need to establish the notation used. Let $\mathbf{X}_{n \times p}$ be the data matrix, with $n$ observations in $p$ dimensions. A d-dimensional projection can be seen as a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$, and defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. Define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ (corresponding to an associated projection matrix $\mathbf{A}$) onto an index value $I$. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a projection.

@@

As a general visualisation method, a tour produces animations of high dimensional data via rotations between low dimension planes. Different tour types choose these planes differently, for example, a grand tour [@cook2008grand] selects the planes randomly and a manual tour [@cook1997manual] gradually phases in and out one variable, to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses planes with the aid of projection pursuit. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs the geodesic interpolation between these planes to form a tour path.

@@

where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthogonal matrices.

Several features of this optimisation are worth noticing. First of all, this is a constrained optimisation problem as the decision variables form the entries of a projection basis, which is required to be orthonormal. It is also likely that the objective function may not be differentiable for a constructed index function and in these cases, gradient-based methods may not work well. Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. Lastly, there is also one computational consideration: the optimisation procedure needs to be fast to compute since the tour animation is played in real-time.

@@

Simulated annealing (`search_better_random`) [@kirkpatrick1983optimization, @bertsimas1993simulated] uses the same sampling process as `search_better` but allows a probabilistic acceptance of a basis with lower index value based on the annealing $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maxima are present. The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

@cook1995grand used a gradient ascent algorithm on the space of the projection bases. In gradient ascent, one first finds the direction for improvement via computing the gradient information. In `search_geodesic`, $2n$ bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter $\delta$. A geodesic is then constructed using the current basis and the one in $2n$ bases with the highest index value. If the neighbourhood parameter $\delta$ is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and works as the searching direction. The next step in gradient ascent is to conduct a line search to find the best improvement along the gradient direction and in `search_geodesic`, this is replaced by optimising the index value along the geodesic direction over an 90 degree angle from $-\pi/4$ to $\pi/4$. The optima $\mathbf{A}_{**}$ is returned for the current iteration if it meets the termination condition on percentage improvement.  The procedure will also terminate if $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. Possible values for `search_better` and `search_better_random` include `random_search`,  `new_basis`, and `interpolation`. `search_geodesic` has a wider variety that includes `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

Multiple iterators are also  needed to index the data collected at different levels. $t$ is a unique identifier that prescribes the natural ordering of each observation; $j$ is the counter for each search-and-interpolate iteration, which remains the same within one round and has an increment of one once a new round starts. $l$ is the counter for each search/interpolation, which provides the information of how many bases the algorithm has searched before finding one to return. There are other parameters of interest, which depend on the particular problem content and they are denoted as *$V_{p}$*. Two most common examples include $V_1 = \text{method}$, which tags the name of the algorithm used, and $V_2 = \text{alpha}$, the neighbourhood parameter that controls the size in sampling candidate bases.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the fact that it is the last iteration means that the optimiser cannot find a better basis and the algorithm terminates. In this notation, the final basis found is $A_T$ with the highest index value $I_T$.

@@

The data structure constructed above meets the tidy data principle [@wickham2014tidy] that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by well-developed packages such as dplyr[@dplyr] and ggplot2[@ggplot2].

The construction of diagnostic plots uses the concept of grammar of graphics [@wickham2010layered] in ggplot2. In grammar of graphics, plots are not defined by their appearance (i.e. boxplot, histogram, scatter plot etc) but by stacked layers. In the construction of diagnostic plots, there are multiple elements one may want to emphasize and there is no single plot name that would meet the need. The stacked layers concept, on the other hand, allows information to be overlaid on each other and one can build the plot from scratch as long as the variables have been stored in a dataset.

## Checking how hard the optimiser is working

A primary interest of diagnosing an optimiser is to study how it progressively finds its optimum. Directly plotting the index value across its natural order will cause the graph to be disproportional to the iteration since it usually takes longer for an optimiser to find a better basis towards the end. Another option is to use summarisation for each iteration. Boxplot is a suitable candidate that can provide five points summary of each iteration. Other additional information not presented in the  boxplot can then be added with new layers, for example text information on the number of points can be added at the bottom of each iteration and the position of basis returned by projection pursuit can be highlighted in point. Further, an option to switch between displaying points and boxplot geometry is helpful when the number of observation is small in one iteration and this is achieved via a `cutoff` parameter.

Figure \@ref(fig:toy-search) shows a sample of the plot constructed using `search_better` with different values of the parameter `max_tries`. Comparing the returned index value of each iteration shows that the termination at `max_tries = 25` is not sufficient for `search_better` to explore the parameter space and a value of 500 is preferred over 25 in this context.

@@

(ref:toy-search) A comparison of `search_better` with different values of the parameter `max_tries`. A six-variable dataset `boa6` is used with `holes` index on a 2D problem. `max_tries = 500` allows the optimiser to find better basis with higher index value at iteration six.

@@

Points on the interpolation path are another interest in tour since the projection on these bases will be played by the tour animation. Figure \@ref(fig:toy-interp) presents the interpolation of three different tour paths each with different curvature. The leftmost plot shows an interpolation where the index value increases progressively and monotonically. The middle path has an increase-then-decrease pattern in the last two iterations and the rightmost path shows even a decrease in the index value at iteration three. The middle situation can be avoid via a construction of the interruption, which will be detailed in section \ref{monotonic} and the rightmost case is a deliberate construction of `search_better_random` where an inferior basis can be accepted with some probability so as to avoid being trapped in a local maximum.

@@

(ref:toy-pca) 1D projection on the 5-variable dataset `boa5` with two optimisers: `search_better` and `search_geodesic`. The yellow point corresponds to the theoretical best basis [0, 1, 0, 0, 0] with `V2` being the only non-normal variable in the dataset.  The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in `search_geodesic` has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases $\mathbf{A}_{p \times d}$ live on the surface of a $p \times d$ dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space ($d > 3$), an easier alternative is to view the reduced space via some dimension reduction methods, i.e. principal component analysis. To better perceive the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package `geozoo` and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \@ref(fig:toy-pca) plots the first two principal components of two search paths, one using `search_better` and the other using `search_geodesic`. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in `search_better` is broader, controlled by `alpha` parameter, which is default to 0.5 while the directional search in `search_geodesic`, controlled by `delta` with a default of 0.01, is so tiny that it can barely be seen.

@@

(ref:toy-tour) A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original $p \times d$ dimension enables a more holistic stereoscopic view of the search. To view a high dimensional ($d \ge 3$) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \@ref(fig:toy-tour) gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \@ref(fig:toy-tour) is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \@ref(fig:toy-pca).

@@

For a particular index function, the best algorithm to optimise relates to the character of the index and the data. If the index function is smooth and has a single maximum, all of the three algorithms introduced above can find the maximum.  When multiple optima are presented, `search_better` may stuck at the local maximum and in the case where the index function is non-smooth, `search_geodesic` may even fail to find the maximum. In this section, examples will be presented  to outline how the diagnostic plots can be used to compare the performance of optimisers in different scenarios.
